---
title: "Simulation Challenge"
subtitle: "Generative Models and Monte Carlo Simulation"
format:
  html:
    theme: cosmo
    toc: false
    number-sections: false
    code-fold: false
    code-tools: false
    css: |
      .cell-output {
        background-color: #f8f9fa;
        border-left: 4px solid #007bff;
        padding: 15px;
        margin: 10px 0;
        border-radius: 5px;
      }
      .cell-output pre {
        background-color: transparent;
        border: none;
        margin: 0;
        padding: 0;
      }
execute:
  echo: false
  eval: true
  warning: false
  message: false
---

# Simulation Challenge - Monte Carlo Analysis

```{r}
#| label: complete-simulation

# Load required libraries
library(dplyr)
library(ggplot2)
library(purrr)
library(tibble)
library(scales)

# Set global options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Game parameters
initial_balance <- 1000
heads_multiplier <- 1.5  # +50%
tails_multiplier <- 0.6  # -40%
years_to_retirement <- 35  # Assuming starting at age 20
num_simulations <- 10000

# Function to simulate a single game
simulate_game <- function(initial_balance, years, heads_mult, tails_mult) {
  balance <- initial_balance
  
  # Track the balance over time for this simulation
  balance_history <- numeric(years + 1)
  balance_history[1] <- balance
  
  for (year in 1:years) {
    # Flip coin (0 = tails, 1 = heads)
    coin_flip <- rbinom(1, 1, 0.5)
    
    if (coin_flip == 1) {
      # Heads: increase by 50%
      balance <- balance * heads_mult
    } else {
      # Tails: decrease by 40%
      balance <- balance * tails_mult
    }
    
    balance_history[year + 1] <- balance
  }
  
  return(list(
    final_balance = balance,
    balance_history = balance_history
  ))
}

# Run many simulations
set.seed(20250923)  # For reproducibility

```

```{r}
#| include: false

# Ensure expected value variables exist for downstream analysis
prob_heads <- 0.5
prob_tails <- 0.5
heads_outcome <- initial_balance * heads_multiplier
tails_outcome <- initial_balance * tails_multiplier
expected_value_1_flip <- prob_heads * heads_outcome + prob_tails * tails_outcome

invisible({
  ev_analysis <- list(
    initial_balance = initial_balance,
    heads_outcome = heads_outcome,
    tails_outcome = tails_outcome,
    expected_value_1_flip = expected_value_1_flip,
    expected_return_pct = (expected_value_1_flip / initial_balance - 1) * 100
  )
})
```

## Part 1: Expected Value Analysis

The expected value after one coin flip is:

**Heads outcome:** $`r format(ev_analysis$heads_outcome, big.mark = ",")` (50% chance)

**Tails outcome:** $`r format(ev_analysis$tails_outcome, big.mark = ",")` (50% chance)

**Expected value:** $`r format(ev_analysis$expected_value_1_flip, big.mark = ",")`

## Part 2: Expectation vs. Reality

```{r}
#| label: expectation-vs-reality

# Calculate difference
difference <- expected_value_1_flip - initial_balance
is_positive <- expected_value_1_flip > initial_balance

# Store for display
invisible({
  expectation_analysis <- list(
    expected_value_1_flip = expected_value_1_flip,
    initial_balance = initial_balance,
    difference = difference,
    is_positive = is_positive,
    expected_return_pct = (expected_value_1_flip / initial_balance - 1) * 100
  )
})
```

**Analysis:**

- **Expected value after 1 flip:** $`r format(expectation_analysis$expected_value_1_flip, big.mark = ",")`
- **Initial investment:** $`r format(expectation_analysis$initial_balance, big.mark = ",")`
- **Difference:** $`r format(expectation_analysis$difference, big.mark = ",")`

`r if(expectation_analysis$is_positive) {" **POSITIVE EXPECTED VALUE:** The expected value is HIGHER than the initial investment."} else {" **NEGATIVE EXPECTED VALUE:** The expected value is LOWER than the initial investment."}`

This means the game has a `r if(expectation_analysis$is_positive) {"positive"} else {"negative"}` expected return of `r round(expectation_analysis$expected_return_pct, 1)`% per flip.

Based on this, you would expect your account to be worth `r if(expectation_analysis$is_positive) {"MORE"} else {"LESS"}` than $1,000.


## Part 3: Single Simulation

```{r}
#| label: single-simulation

# Run a single simulation and track balance over time
single_sim <- simulate_game(initial_balance, years_to_retirement, heads_multiplier, tails_multiplier)

# Create data for plotting
years <- 0:years_to_retirement
single_sim_data <- tibble(
  year = years,
  age = 20 + years,  # Starting at age 20
  balance = single_sim$balance_history,
  return_pct = (balance / initial_balance - 1) * 100
)

# Create the plot
p_single <- ggplot(single_sim_data, aes(x = age, y = balance)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "darkblue", size = 1) +
  geom_hline(yintercept = initial_balance, color = "red", linetype = "dashed", size = 1) +
  scale_y_continuous(labels = dollar_format(), trans = "log10") +
  labs(
    title = "Single Simulation: Account Balance Over Time",
    subtitle = paste("Final balance: $", format(round(single_sim$final_balance, 2), big.mark = ","), 
                     " (", round((single_sim$final_balance / initial_balance - 1) * 100, 1), "% return)"),
    x = "Age",
    y = "Account Balance ($, Log Scale)",
    caption = "Red dashed line = Initial $1,000 investment"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 14, color = "darkblue"),
    plot.caption = element_text(size = 10, color = "gray"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

print(p_single)

# Analysis of this single simulation
estimated_heads <- (log(single_sim$final_balance / initial_balance) - years_to_retirement * log(0.6)) / (log(1.5) - log(0.6))
estimated_tails <- years_to_retirement - estimated_heads
total_return_pct <- (single_sim$final_balance / initial_balance - 1) * 100
annualized_return <- (single_sim$final_balance / initial_balance)^(1/years_to_retirement) - 1
is_profitable <- single_sim$final_balance > initial_balance

# Store for display
invisible({
  single_sim_analysis <- list(
    starting_age = 20,
    retirement_age = 55,
    initial_balance = initial_balance,
    final_balance = single_sim$final_balance,
    total_return_pct = total_return_pct,
    annualized_return = annualized_return,
    estimated_heads = estimated_heads,
    estimated_tails = estimated_tails,
    is_profitable = is_profitable
  )
})
```

**Single Simulation Analysis:**

- **Starting age:** `r single_sim_analysis$starting_age`
- **Retirement age:** `r single_sim_analysis$retirement_age`
- **Initial balance:** $`r format(single_sim_analysis$initial_balance, big.mark = ",")`
- **Final balance:** $`r format(round(single_sim_analysis$final_balance, 2), big.mark = ",")`
- **Total return:** `r round(single_sim_analysis$total_return_pct, 1)`%
- **Annualized return:** `r round(single_sim_analysis$annualized_return * 100, 3)`%
- **Estimated heads:** `r round(single_sim_analysis$estimated_heads, 1)` out of `r years_to_retirement` flips
- **Estimated tails:** `r round(single_sim_analysis$estimated_tails, 1)` out of `r years_to_retirement` flips

**Commentary:**

`r if(single_sim_analysis$is_profitable) {"**SUCCESS!** This simulation resulted in a **PROFIT**."} else {"**LOSS!** This simulation resulted in a **LOSS**."}`

The account `r if(single_sim_analysis$is_profitable) {"grew"} else {"shrank"}` from $1,000 to $`r format(round(single_sim_analysis$final_balance, 2), big.mark = ",")`.

This represents a `r round(single_sim_analysis$total_return_pct, 1)`% `r if(single_sim_analysis$is_profitable) {"gain"} else {"loss"}` over 35 years.

This single simulation shows the high volatility of the game. While the expected value is positive, individual outcomes can vary dramatically. I would not be happy with a 98.4% loss on my investment after 35 years of the game.

## Part 4: Multiple Simulations
```{r}
#| label: multiple-simulations

# Run 100 simulations
set.seed(20250924)  # Different seed for variety
num_small_simulations <- 100

# Run 100 simulations
small_simulation_results <- map(1:num_small_simulations, function(i) {
  simulate_game(initial_balance, years_to_retirement, heads_multiplier, tails_multiplier)
})

# Extract final balances
small_final_balances <- map_dbl(small_simulation_results, ~ .x$final_balance)

# Create results dataframe
small_results_df <- tibble(
  simulation = 1:num_small_simulations,
  final_balance = small_final_balances,
  return_pct = (small_final_balances / initial_balance - 1) * 100
)

# Calculate summary statistics for 100 simulations
small_summary_stats <- small_results_df %>%
  summarise(
    mean_balance = mean(final_balance),
    median_balance = median(final_balance),
    min_balance = min(final_balance),
    max_balance = max(final_balance),
    std_dev = sd(final_balance),
    prob_positive = mean(return_pct > 0),
    prob_break_even = mean(final_balance >= initial_balance)
  )

# Create probability distribution plot
p_distribution <- ggplot(small_results_df, aes(x = final_balance)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, color = "black", 
                 aes(y = after_stat(density))) +
  geom_density(color = "red", size = 1.2, alpha = 0.8) +
  geom_vline(xintercept = initial_balance, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = small_summary_stats$mean_balance, color = "green", linetype = "dashed", size = 1) +
  geom_vline(xintercept = small_summary_stats$median_balance, color = "orange", linetype = "dashed", size = 1) +
  scale_x_continuous(labels = dollar_format(), trans = "log10") +
  labs(
    title = "Probability Distribution of Final Balances (100 Simulations)",
    subtitle = paste("Mean: $", format(round(small_summary_stats$mean_balance, 2), big.mark = ","),
                     " | Median: $", format(round(small_summary_stats$median_balance, 2), big.mark = ","),
                     " | Std Dev: $", format(round(small_summary_stats$std_dev, 2), big.mark = ",")),
    x = "Final Balance at Age 55 ($, Log Scale)",
    y = "Probability Density",
    caption = "Red line = Initial $1,000 | Green line = Mean | Orange line = Median"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "darkblue"),
    plot.caption = element_text(size = 10, color = "gray"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

print(p_distribution)

# Store results for display
invisible({
  small_sim_analysis <- list(
    num_simulations = num_small_simulations,
    mean_balance = small_summary_stats$mean_balance,
    median_balance = small_summary_stats$median_balance,
    min_balance = small_summary_stats$min_balance,
    max_balance = small_summary_stats$max_balance,
    std_dev = small_summary_stats$std_dev,
    prob_positive = small_summary_stats$prob_positive,
    prob_break_even = small_summary_stats$prob_break_even
  )
})
```

**100 Simulations Summary:**

```{r}
#| label: small-summary-table
#| results: asis

# Compute local probabilities to avoid chunk order dependencies
prob_gt_1000_local <- mean(small_final_balances > initial_balance)
prob_ge_1000_local <- mean(small_final_balances >= initial_balance)

small_summary_table <- tibble::tibble(
  Metric = c(
    "Number of simulations",
    "Mean final balance",
    "Median final balance",
    "Minimum final balance",
    "Maximum final balance",
    "Standard deviation",
    "Probability > $1,000"
  ),
  Value = c(
    format(small_sim_analysis$num_simulations, big.mark = ","),
    scales::dollar(round(small_sim_analysis$mean_balance, 2)),
    scales::dollar(round(small_sim_analysis$median_balance, 2)),
    scales::dollar(round(small_sim_analysis$min_balance, 2)),
    scales::dollar(round(small_sim_analysis$max_balance, 2)),
    scales::dollar(round(small_sim_analysis$std_dev, 2)),
    paste0(round(prob_gt_1000_local * 100, 1), "%")
  )
)

knitr::kable(small_summary_table, align = c('l','r'), caption = 'Summary Statistics (100 Simulations)')
```

**Key Observations:**

The probability distribution reveals the **extreme volatility** of this investment game:

- **Wide range:** From $`r format(round(small_sim_analysis$min_balance, 2), big.mark = ",")` to $`r format(round(small_sim_analysis$max_balance, 2), big.mark = ",")`
- **High standard deviation:** $`r format(round(small_sim_analysis$std_dev, 2), big.mark = ",")` indicates massive variability
- **Skewed distribution:** The mean ($`r format(round(small_sim_analysis$mean_balance, 2), big.mark = ",")`) is much higher than the median ($`r format(round(small_sim_analysis$median_balance, 2), big.mark = ",")`)
- **Most outcomes are losses:** Only `r round(small_sim_analysis$prob_positive * 100, 1)`% of simulations resulted in positive returns

As an investor, I would not be happy playing this game long term because almost 70% of the time I would end up with less than my initial investment. There are a few outliers that skew the average but the median of $250 is a good indicator of the typical outcome.

## Part 5: Probability Analysis

The probability that my account balance will be greater than $1,000 at age 55 is 31%.

```{r}
#| label: probability-analysis

# Calculate probability of ending with more than $1,000
prob_greater_than_1000 <- mean(small_final_balances > initial_balance)
prob_less_than_1000 <- mean(small_final_balances < initial_balance)

# Calculate additional probability thresholds
prob_greater_than_2000 <- mean(small_final_balances > 2000)
prob_greater_than_5000 <- mean(small_final_balances > 5000)
prob_greater_than_10000 <- mean(small_final_balances > 10000)

# Calculate percentiles
percentiles <- quantile(small_final_balances, probs = c(0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99))

# Calculate percentile where value equals $1000 (interpolation)
percentile_1000 <- ecdf(small_final_balances)(1000) * 100

# Store results for display
invisible({
  probability_analysis <- list(
    prob_greater_than_1000 = prob_greater_than_1000,
    prob_less_than_1000 = prob_less_than_1000,
    prob_greater_than_2000 = prob_greater_than_2000,
    prob_greater_than_5000 = prob_greater_than_5000,
    prob_greater_than_10000 = prob_greater_than_10000,
    percentiles = percentiles,
    percentile_1000 = percentile_1000
  )
})
```

**Probability Analysis Results:**

```{r}
#| label: probability-table
#| results: asis

pct_fmt <- scales::dollar_format(accuracy = 0.01)

pct_table <- tibble::tibble(
  `Percentile` = c(
    "10th",
    "25th",
    "50th (median)",
    paste0(round(probability_analysis$percentile_1000, 1), "th ($1,000)"),
    "75th",
    "90th",
    "95th",
    "99th"
  ),
  `Value` = c(
    pct_fmt(probability_analysis$percentiles["10%"]),
    pct_fmt(probability_analysis$percentiles["25%"]),
    pct_fmt(probability_analysis$percentiles["50%"]),
    "$1,000.00",
    pct_fmt(probability_analysis$percentiles["75%"]),
    pct_fmt(probability_analysis$percentiles["90%"]),
    pct_fmt(probability_analysis$percentiles["95%"]),
    pct_fmt(probability_analysis$percentiles["99%"])
  )
)

knitr::kable(pct_table, align = c('l','r'), caption = 'Percentile Summary (100 Simulations)')
```

## Part 6: Strategy Comparison

```{r}
#| label: strategy-comparison
#| results: asis

# Modified game logic:
# Bet exactly 50% of balance each year; the other 50% remains unchanged.
# Effective annual multipliers: Heads -> 0.5 + 0.5*1.5 = 1.25; Tails -> 0.5 + 0.5*0.6 = 0.8
mod_heads_multiplier <- 1.25
mod_tails_multiplier <- 0.8

set.seed(20250925)
num_comparison_sims <- 100

mod_results <- purrr::map(1:num_comparison_sims, function(i) {
  simulate_game(initial_balance, years_to_retirement, mod_heads_multiplier, mod_tails_multiplier)
})

mod_final_balances <- purrr::map_dbl(mod_results, ~ .x$final_balance)

# Probabilities > $10,000
prob_mod_gt_10000 <- mean(mod_final_balances > 10000)

# Use original 100-sim results from Section 4 (small_final_balances)
prob_orig_gt_10000 <- mean(small_final_balances > 10000)

comparison_word <- if (abs(prob_mod_gt_10000 - prob_orig_gt_10000) < 1e-9) "Equal" else if (prob_mod_gt_10000 > prob_orig_gt_10000) "Higher" else "Lower"

comparison_table <- tibble::tibble(
  Scenario = c("Original game", "Modified"),
  `P(Final > $10,000)` = c(
    paste0(round(prob_orig_gt_10000 * 100, 1), "%"),
    paste0(round(prob_mod_gt_10000 * 100, 1), "%")
  )
)

cat("\n**Analysis:** The modified game strategy has a", if (prob_mod_gt_10000 > prob_orig_gt_10000) {"higher"} else {"lower"}, "probability of ending with more than $10,000 at age 55 compared to the original game. Interestingly, while the modified strategy is more likely to achieve at least $1,000 (break-even), it is less likely to reach the high-end target of $10,000. This suggests that the modified strategy is", if (prob_mod_gt_10000 > prob_orig_gt_10000) {"more likely to achieve high returns"} else {"less likely to achieve high returns"}, "but may also be", if (prob_mod_gt_10000 > prob_orig_gt_10000) {"more volatile"} else {"less volatile"}, "overall.\n")

knitr::kable(comparison_table, align = c('l','r'), caption = 'Probability of Ending Above $10,000 (100 Simulations)')

cat("\n", sprintf("Result: %s (modified vs. original)", comparison_word))
```

## Visualizations and Tables: Original vs. Modified

```{r}
#| label: comparison-visuals
#| fig-cap: "Distribution of final balances across 100 simulations for both strategies"

# Build combined dataset for visualization
orig_df_small <- tibble::tibble(
  simulation = 1:length(small_final_balances),
  final_balance = small_final_balances,
  scenario = "Original"
)

mod_df_small <- tibble::tibble(
  simulation = 1:length(mod_final_balances),
  final_balance = mod_final_balances,
  scenario = "Modified"
)

combined_df <- dplyr::bind_rows(orig_df_small, mod_df_small)

# Histogram + density by scenario (faceted)
p_comp <- ggplot(combined_df, aes(x = final_balance, fill = scenario)) +
  geom_histogram(bins = 20, alpha = 0.6, color = "black", aes(y = after_stat(density))) +
  geom_density(alpha = 0.4) +
  scale_x_continuous(labels = scales::dollar, trans = "log10") +
  labs(
    title = "Final Balance Distributions (Log Scale)",
    x = "Final Balance at Age 55 ($, Log Scale)",
    y = "Density",
    fill = "Scenario"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom"
  ) +
  facet_wrap(~ scenario, ncol = 1, scales = "free_y")

print(p_comp)
```

```{r}
#| label: comparison-box
#| fig-cap: "Boxplots of final balances across 100 simulations by strategy"

ggplot(combined_df, aes(x = scenario, y = final_balance, fill = scenario)) +
  geom_boxplot(alpha = 0.6, outlier.alpha = 0.5) +
  scale_y_continuous(labels = scales::dollar, trans = "log10") +
  labs(
    title = "Final Balance Boxplots (Log Scale)",
    x = "Strategy",
    y = "Final Balance ($)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "none"
  )
```

```{r}
#| label: comparison-tables
#| results: asis

summary_compare <- combined_df %>%
  group_by(scenario) %>%
  summarise(
    mean_balance = mean(final_balance),
    median_balance = median(final_balance),
    min_balance = min(final_balance),
    max_balance = max(final_balance),
    sd_balance = sd(final_balance),
    prob_gt_1000 = mean(final_balance > 1000),
    prob_gt_10000 = mean(final_balance > 10000),
    .groups = 'drop'
  ) %>%
  mutate(
    mean_balance = scales::dollar(round(mean_balance, 2)),
    median_balance = scales::dollar(round(median_balance, 2)),
    min_balance = scales::dollar(round(min_balance, 2)),
    max_balance = scales::dollar(round(max_balance, 2)),
    sd_balance = scales::dollar(round(sd_balance, 2)),
    `P(> $1,000)` = paste0(round(prob_gt_1000 * 100, 1), "%"),
    `P(> $10,000)` = paste0(round(prob_gt_10000 * 100, 1), "%")
  ) %>%
  select(Scenario = scenario, `Mean` = mean_balance, `Median` = median_balance, `Min` = min_balance,
         `Max` = max_balance, `Std Dev` = sd_balance, `P(> $1,000)`, `P(> $10,000)`)

knitr::kable(summary_compare, align = c('l','r','r','r','r','r','r','r'), caption = 'Summary Comparison: Original vs Modified (100 Simulations)')
```

---