---
title: "Simulation Challenge"
subtitle: "Generative Models and Monte Carlo Simulation"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This document presents a simulation challenge focused on generative models and Monte Carlo simulation techniques.

## Objectives

- Implement Monte Carlo simulation methods
- Explore generative modeling approaches
- Analyze simulation results and convergence
- Visualize simulation outcomes

# Setup

## Load Required Libraries

```{r load-libraries}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(reshape2)
library(gridExtra)
```

# Monte Carlo Simulation

## Basic Monte Carlo Example

```{r basic-monte-carlo}
# Simple Monte Carlo estimation of π
monte_carlo_pi <- function(n) {
  x <- runif(n, -1, 1)
  y <- runif(n, -1, 1)
  inside <- sum(x^2 + y^2 <= 1)
  return(4 * inside / n)
}

# Run simulation
set.seed(123)
n_sim <- 10000
pi_estimate <- monte_carlo_pi(n_sim)
cat("Estimated π:", pi_estimate, "\n")
cat("Actual π:", pi, "\n")
cat("Error:", abs(pi_estimate - pi), "\n")
```

## Visualization

```{r pi-visualization, fig.width=10, fig.height=5}
# Create visualization
set.seed(123)
n_vis <- 1000
x <- runif(n_vis, -1, 1)
y <- runif(n_vis, -1, 1)
inside <- x^2 + y^2 <= 1

df <- data.frame(x = x, y = y, inside = inside)

p1 <- ggplot(df, aes(x = x, y = y, color = inside)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("FALSE" = "red", "TRUE" = "blue")) +
  labs(title = "Monte Carlo Estimation of π",
       subtitle = paste("Points inside circle:", sum(inside), "out of", n_vis),
       x = "X", y = "Y") +
  theme_minimal()

# Convergence plot
n_values <- seq(100, 10000, by = 100)
pi_estimates <- sapply(n_values, monte_carlo_pi)

p2 <- ggplot(data.frame(n = n_values, pi_est = pi_estimates), 
             aes(x = n, y = pi_est)) +
  geom_line() +
  geom_hline(yintercept = pi, linetype = "dashed", color = "red") +
  labs(title = "Convergence of π Estimation",
       x = "Number of Simulations", y = "Estimated π") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

# Generative Models

## Simple Generative Model

```{r generative-model}
# Generate synthetic data using a simple generative model
generate_data <- function(n, mean = 0, sd = 1) {
  return(rnorm(n, mean, sd))
}

# Generate training data
set.seed(123)
train_data <- generate_data(1000, mean = 5, sd = 2)

# Visualize generated data
ggplot(data.frame(x = train_data), aes(x = x)) +
  geom_histogram(bins = 30, alpha = 0.7, fill = "skyblue") +
  labs(title = "Generated Data Distribution",
       x = "Value", y = "Frequency") +
  theme_minimal()
```

# Results and Analysis

## Simulation Results

The Monte Carlo simulation successfully estimated π with reasonable accuracy. The convergence plot shows how the estimate improves with more simulations.

## Key Findings

- Monte Carlo methods provide good approximations for complex integrals
- Convergence improves with sample size
- Visual inspection helps validate simulation results

# Conclusion

This simulation challenge demonstrates the power of Monte Carlo methods and generative models for statistical estimation and data generation.

# References

- Monte Carlo Methods in Statistical Computing
- Introduction to Generative Models
- R for Data Science